

%
%  HBCmain
%
%  Created by Steven James Dean Martell on 2011-09-27.
%  Copyright (c) 2011 UBC Fisheries Centre. All rights reserved.
%
\documentclass[12pt]{article}

% Use utf-8 encoding for foreign characters
\usepackage[utf8]{inputenc}

% Setup for fullpage use
\usepackage{fullpage}
\usepackage{url}

% Uncomment some of the following if you use the features
%
% Running Headers and footers
%\usepackage{fancyhdr}
% Multipart figures
%\usepackage{subfigure}
% Natbib package for bibliography
\usepackage[round]{natbib}

% More symbols
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{latexsym}

% Surround parts of graphics with box
\usepackage{boxedminipage}

% Package for including code in the document
\usepackage{listings}

% If you want to generate a toc for each chapter (use with book)
\usepackage{minitoc}

% This is now the recommended way for checking for PDFLaTeX:
\usepackage{ifpdf}

%% The following is used for tables of equations.
\newcounter{saveEq} 
\def\putEq{\setcounter{saveEq}{\value{equation}}} 
\def\getEq{\setcounter{equation}{\value{saveEq}}} 
\def 
\tableEq{ 

% equations in tables
\putEq \setcounter{equation}{0} 
\renewcommand{\theequation}{T\arabic{table}.\arabic{equation}} \vspace{-5mm} } 
\def\normalEq{ 

% renew normal equations
\getEq 
\renewcommand{\theequation}{\arabic{section}.\arabic{equation}}}

\def\puthrule{ 

%thick rule lines for equation tables
\hrule \hrule \hrule \hrule \hrule} 
\usepackage{multicol}

%\newif\ifpdf
%\ifx\pdfoutput\undefined
%\pdffalse % we are not running PDFLaTeX
%\else
%\pdfoutput=1 % we are running PDFLaTeX
%\pdftrue
%\fi
\ifpdf 
\usepackage[pdftex]{graphicx} \else 
\usepackage{graphicx} \fi 
\title{Estimates of population status of humpback chub (\textit{Gila cypha}) in the Grand Canyon based on a length-based mark-recapture assessment model.} 
\author{ Steven J. D. Martell\\
UBC Fisheries Centre,\\
2202 Main Mall,\\
Vancouver, BC\\
V6T 1Z4,\\
CANADA }

\date{2011-09-27}

\begin{document}

\ifpdf \DeclareGraphicsExtensions{.pdf, .jpg, .tif} \else \DeclareGraphicsExtensions{.eps, .jpg} \fi

\maketitle 
\begin{abstract}
	A length-structured population model that incorporates mark-recapture data was developed to estimate abundance of humpback chub (\textit{Gila cypha}) in the Grand Canyon between 1989 and 2011. The model was fit to observations on catch-at-length and capture-recapture data from sampling programs that employed a variety of gear types including electrofishing, tramel netting, and hoop nets. Previous assessment models were age-based and have been shown to produce biased estimates of recent recruitment estimates. The age-structured model was also limited to data on fish that were greater than 150mm total length; the minimum size required for tagging. The length-structured model developed here makes no assumptions about age of individual fish and is also fit to catch-at-length data for fish that are too small to tag. Model outputs are based the number of fish greater than a specified size, or in a fixed size interval, there are no age-based results. 
\end{abstract}

\tableofcontents

\section{Introduction}\label{sec:Intro}

In this report, I develop a length-structured mark-recapture model (hereafter, LSMR) where the accounting system for population numbers is based purely on length. The model is a statistical catch-at-length model, where the initial length distribution and recruitment each year are treated as latent variables to be estimated by fitting the model to a series of catch-at-length observations take over a period of time. A separable function is developed to estimate the year and size effect in observed catch-at-length data. The statistical nature of the model is very similar to that of \cite{fournier1982general}, but is based on catch-at-length rather than catch-age data. The model is also fit to length-based capture-recapture data, where the growth and survival of tagged animals is updated at each time step and the predicted ratio of marked and unmarked fish-at-length is used to estimate the length-based recapture rates.

\input{./Methods/LSMRmethods}

\bibliographystyle{apalike} 
\bibliography{$HOME/Documents/ARTICLES/Articles-1}

\appendix 
\include{./Appendix1/WorkPlan} 
\include{./Appendix2/SimulationModel}

\end{document} 
